Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-21 paper/paper.tex
cfgs

\begin{figure}
  \begin{tikzpicture}[node distance = 17mm ]
    \node[state, initial, accepting] (0) {$0$};
    \node[state, right of=0] (1) {$1$};
    \node[state, right of=1] (2) {$2$};
    \node[right of=2] (3) {$\dots$};
    \node[state, below of=0] (fail) {$fail$};

    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
              (1) edge[below, bend left] node{\stringquote{)}} (0)
              (1) edge[above, bend left] node{\stringquote{(}} (2)
              (2) edge[below, bend left] node{\stringquote{)}} (1)
              (2) edge[above, bend left] node{\stringquote{(}} (3)
              (3) edge[below, bend left] node{\stringquote{)}} (2)
              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
              (0) edge[left] node{\stringquote{)}} (fail);
  \end{tikzpicture}
  \caption{Automaton $M$ for the Dyck grammar}
  \label{fig:DyckAutomaton}
\end{figure}

\steven{Maybe this detail should go closer to the actual example
We use distributivity to encode the of the LL(1) parser in
\cref{thm:exp-parser}. At parse time, an LL(1) automaton peeks at the next
character of a string and uses the peeked character to inform its next
transition. Our automaton from \cref{fig:binop-inductive} chooses which state to
transition to dependent on if the next character is $\stringquote{)}$.

TODO

the proof obligation is

$\CharGram \otimes \LinPiTy{s}{state}{Trace s true}$ which is iso to
$(\literal{)} \otimes \top \oplus NotStatsWithRP) \otimes
\LinPiTy{s}{state}{Trace s true}$.

However, without distributivity we get stuck and can go further to actually use
$lookaheadRP$ or $lookaheadNot$.

}

Our final example is of a simple grammar of arithmetic expressions
with an associative operation. Here we take the alphabet to be $\{
\stringquote{(}, \stringquote{)}, \stringquote{+} , \stringquote{NUM}
\}$. In \Cref{fig:binop-inductive} we define it using two mutually
recursive types, corresponding to the two non-terminals we would use
in a CFG syntax. The syntactic structure encodes that the binary
operation is right associative. In the same figure, we define the
traces of an automaton with one token of lookahead. The automaton has
four different ``states'', each with access to a natural number
``stack''. The ``opening'' state $O$ expects either an open paren, in
which case it increments the stack and stays in the opening state, or
sees a number and proceeds to the $D$ state. The ``done opening''
state $D$ is where lookahead is used: if the next token will be a
right paren, then we proceed to $C$, otherwise we proceed to $M$. Here
$\mathsf{NotStartsWithRP}$ is defined as
$\I \oplus (({\literal{(}} \oplus {\literal {+}}
\oplus \mathsf{NUM}) \otimes \top)$. In the ``closing'' state $C$ if
we observe a close paren we decrement the count and continue to the
$D$ state. In the ``multiplying'' state $M$, we succeed if the string
ends and the count is $0$, and if we see a plus we continue to the $O$
state. Additionally, since the automaton need also parse all of the
incorrect strings, we add additional failing cases. It is
straightforward to implement a parser for this lookahead automaton,
generalizing the approach for deterministic automata.

\begin{theorem}[\agdalogo]
  \label{thm:exp-parser}
  We construct a parser for $\mathsf{Exp}$ by showing it is weakly
  equivalent to $\mathsf{O}~0~\true$.
\end{theorem}
\begin{figure}
\begin{lstlisting}
data Exp : L where
  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
data Atom : L where
  num : (*@$\uparrow$@*)(NUM (*@$\lto$@*) Atom)
  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)

data O : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) DO n b (*@$\lto$@*) O n b)
  done : (*@$\uparrow$@*)(&[ n ] O n false)
  unexpected : (*@$\uparrow$@*)(&[ n ] (')' (*@$\oplus$@*) '+') (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) O n false)
data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  lookAheadRP  : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
  lookAheadNot  : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) M n b) (*@$\lto$@*) D n b)
data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
  closeBad : (*@$\uparrow$@*)(&[ n ] ')' (*@$\lto$@*) C 0 b)
  done : (*@$\uparrow$@*)(&[ n ] C n false)
  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) '+' (*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) C n false)
data M : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  doneGood : (*@$\uparrow$@*)(M 0 true)
  doneBad : (*@$\uparrow$@*)(&[ n ] M (n + 1) false)
  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) M n b)
  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) M n false)
\end{lstlisting}
\caption{Associative arithmetic expressions and a corresponding lookahead automaton}
\label{fig:binop-inductive}
\end{figure}
% \subsection{LL(1)}
% %% 4. LL(1) parser

>>>>

\begin{figure}
  \begin{tikzpicture}[node distance = 17mm ]
    \node[state, initial above, accepting] (0) {$0$};
    \node[state, right of=0] (1) {$1$};
    \node[state, right of=1] (2) {$2$};
    \node[right of=2] (3) {$\dots$};
    \node[state, left of=0] (fail) {$fail$};

    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
              (1) edge[below, bend left] node{\stringquote{)}} (0)
              (1) edge[above, bend left] node{\stringquote{(}} (2)
              (2) edge[below, bend left] node{\stringquote{)}} (1)
              (2) edge[above, bend left] node{\stringquote{(}} (3)
              (3) edge[below, bend left] node{\stringquote{)}} (2)
              (fail) edge[loop left] node{\stringquote{(}, \stringquote{)}} (fail)
              (0) edge[above] node{\stringquote{)}} (fail);
  \end{tikzpicture}
  \caption{Automaton $M$ for the Dyck grammar}
  \label{fig:DyckAutomaton}
\end{figure}

Our final example is of a simple grammar of arithmetic expressions
with an associative operation. Here we take the alphabet to be $\{
\stringquote{(}, \stringquote{)}, \stringquote{+} , \stringquote{NUM}
\}$. In \Cref{fig:binop-inductive} we define the expression grammar
using two mutually
recursive types, corresponding to the two non-terminals we would use
in a CFG syntax. The syntactic structure encodes that the binary
operation is right associative. In the same figure, we define the
traces of an automaton with one token of lookahead. The automaton\footnote{In
  the automaton definition,
  $\texttt{NotStartsWithLP}$ is defined as
  \(I \oplus
  (\literal{)} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
  \). Similarly,
  $\texttt{NotStartsWithRP}$ is defined as
  \(I \oplus
  (\literal{(} \oplus \literal{+} \oplus \mathtt{NUM}) \otimes \top
  \).
} has
four different ``states'', each with access to a natural number
``stack''. The ``opening'' state $O$ expects either an left paren, in
which case it increments the stack and stays in the opening state, or
sees a number and proceeds to the $D$ state. The ``done opening''
state $D$ is where lookahead is used: if the next token will be a
right paren, then we proceed to $C$; otherwise, we proceed to $A$.
In the ``closing'' state $C$ if
we observe a right paren, then we decrement the count and continue to the
$D$ state. In the ``adding'' state $A$, we succeed if the string
ends and the count is $0$; otherwise, if we see a plus we continue to the $O$
state. Additionally, since the automaton need parse all of the
incorrect strings, we add all of the failing cases.

\begin{theorem}[\agdalogo]
  \label{thm:exp-parser}
  We construct a parser for $\mathsf{Exp}$ by showing it is weakly
  equivalent to $\mathsf{O}~0~\true$.
\end{theorem}

\steven{
With \cref{ax:dist}, it is straightforward to implement a parser for this lookahead automaton,
generalizing the approach for deterministic automata. Without access to
distributivity, we may define a rudimentary lookahead operation via the chain of
equivalences,

\[
  A~n~b \cong (A~n~b) \& \top \cong (A~n~b)\,\& \left(\left(\literal{)} \otimes \top \right) \oplus \texttt{NotStartsWithRP}\right)
\]

When defining a parser, if the lookahead character is a right paren we would
like to apply $\texttt{lookAheadRP}$; otherwise, apply $\texttt{lookAheadNot}$.
However, without distributivity we have no means of relating the bit of
information learned by looking ahead to the input $A~n~b$ parse.
That is, to choose which constructor to apply we would need the end of this
chain of equivalences to be a binary sum rather than a binary product,
necessitating the distributivity axiom.
}

\begin{figure}
\begin{lstlisting}
data Exp : L where
  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
data Atom : L where
  num : (*@$\uparrow$@*)(NUM (*@$\lto$@*) Atom)
  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)

data O : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
  unexpected : (*@$\uparrow$@*)(&[ n ] (NotStartsWithLP (*@$\lto$@*) O n false)
data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  lookAheadRP : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
  lookAheadNot : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
  closeBad : (*@$\uparrow$@*)(')' (*@$\lto$@*) C 0 false)
  unexpected : (*@$\uparrow$@*)(&[ n ] NotStartsWithRP (*@$\lto$@*) C n false)
data A : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  doneGood : (*@$\uparrow$@*)(A 0 true)
  doneBad : (*@$\uparrow$@*)(&[ n ] A (n + 1) false)
  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) A n false)
\end{lstlisting}
\caption{Associative arithmetic expressions and a corresponding lookahead automaton}
\label{fig:binop-inductive}
\end{figure}
% \subsection{LL(1)}
% %% 4. LL(1) parser
